A Roadmap for\\HEP Software and Computing R\&D\\for the 2020s

### 2018-10-19 CWP/papers/HSF-CWP-2017-01_roadmap/latex/appendices.tex
[WIP] Finalization of CWP version for CSBS journal (#104)
* Use one column for author list

* CWP: appendices as a separate file, one column in cwp-csbs

* Fix a non itemized entry in the glossary (Concurrency Forum)

% Copyright (C) 2017, The HSF Community White Paper authors, licence CC-BY-4.0.

% CSBS template
\documentclass[twocolumn]{svjour3}          % twocolumn
\smartqed  % flush right qed marks, e.g. at end of proof
\journalname{CSBS}

\usepackage{graphicx}
\usepackage{subcaption}
\captionsetup{compatibility=false}
%\usepackage{subfigure}
\usepackage[titletoc]{appendix}
\usepackage{xspace}
\usepackage{hyperref}

\usepackage[utf8]{inputenc}

\usepackage[style=numeric-comp,sorting=none]{biblatex}
\addbibresource{cwp.bib}
\addbibresource{cwp-chapters.bib}

%% Comment out before final submission
\usepackage{lineno}  % for line numbering during review
\linenumbers

\begin{document}

\noindent
\begin{tabular*}{\linewidth}{lc@{\extracolsep{\fill}}r@{\extracolsep{0pt}}}
 & & HSF-CWP-2017-01 \\
 & & December 15, 2017 \\ % use \date or hardwire e.g. December 15, 2017
 & & \\
\end{tabular*}
\vspace{2.0cm}

\title{A Roadmap for\\HEP Software and Computing R\&D\\for the 2020s}

%\subtitle{Do you have a subtitle?\\ If so, write it here}

%\titlerunning{Short form of title}        % if too long for running head

\author{HEP Software Foundation (authors are listed at the end of this report)}

%\author{First Author         \and
%        Second Author %etc.
%}

%\authorrunning{Short form of author list} % if too long for running head

%\institute{F. Author \at
%              first address \\
%              Tel.: +123-45-678910\\
M              Fax: +123-45-678910\\
%              \email{fauthor@example.com}           %  \\
%             \emph{Present address:} of F. Author  %  if needed
%           \and
%           S. Author \at
%              second address
%}

\date{Received: date / Accepted: date}

\maketitle

\abstract{Particle physics has an ambitious and broad experimental
programme for the coming decades. This programme requires large
investments in detector hardware, either to build new facilities
and experiments, or to upgrade existing ones.
Similarly, it requires commensurate investment in the R\&D of software to acquire, manage, process,
and analyse the shear amounts of data to be recorded.
In planning for the HL-LHC in particular, it is critical that all of the
collaborating stakeholders agree on the software goals and priorities,
and that the efforts complement each other. In this spirit, this white paper
describes the R\&D activities required to prepare for this software upgrade.}

\newpage

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

Particle physics has an ambitious experimental programme for the
coming decades. The programme supports the strategic goals of the
particle physics community that have been laid out by the European
Strategy for Particle Physics~\cite{ESPP2013} and by
the Particle Physics Project Prioritization Panel (P5)~\cite{DOE-P5} in the United
States~\cite{2014pwa}. Broadly speaking, the
scientific goals are:

\begin{itemize}
\item Exploit the discovery of the Higgs boson as a precision tool
  for investigating Standard Model (SM) and Beyond the Standard Model
  (BSM) physics.
\item
  Study the decays of $b$- and $c$-hadrons, and tau leptons, in the search
  for manifestations of BSM physics, and investigate
  matter-antimatter differences.
\item
  Search for signatures of dark matter.
\item
  Probe neutrino oscillations and masses.
\item
  Study the Quark Gluon Plasma state of matter in heavy-ion collisions.
\item
  Explore the unknown.
\end{itemize}

The High-Luminosity Large Hadron Collider
(HL-LHC)~\cite{HL-LHC,1742-6596-515-1-012012,Apollinari:2284929} 
will be a major
upgrade of the current LHC~\cite{LHC} supporting the aim of 
an in-depth investigation of the properties of the Higgs boson and its
couplings to other particles (Figure~\ref{fig:hl-lhc}). The ATLAS~\cite{ATLAS} and CMS~\cite{CMS}
collaborations will continue to make measurements in the Higgs sector,
while searching for new physics Beyond the Standard Model (BSM). Should
a BSM discovery be made, a full exploration of that physics will be
pursued. Such BSM physics may help shed light on the nature of dark
matter, which we know makes up the majority of gravitational matter in
the universe, but which does not interact via the electromagnetic or
strong nuclear forces~\cite{Mangano2016}.

The LHCb experiment at the LHC~\cite{LHCb} and the Belle II experiment at
KEK~\cite{BelleII} study various aspects of heavy
flavour physics (b- and c-quark, and tau-lepton physics), where quantum
influences of very high mass particles manifest themselves in lower
energy phenomena. Their primary goal is to look for BSM physics, either
by studying CP violation (that is, asymmetries in the behaviour of
particles and their corresponding antiparticles) or modifications in
rate or angular distributions in rare heavy-flavour decays. Current
manifestations of such asymmetries do not explain why our universe is so
matter dominated. These flavour physics programmes are related to BSM
searches through effective field theory, and powerful constraints on new
physics keep coming from such studies.

\begin{figure}[bthp]
    \centering
    \includegraphics[width=0.94\textwidth]{HL-LHC-plan-2017-Plan-2.pdf}
    \caption{The current schedule for the LHC and HL-LHC upgrade
and run~\cite{HL-LHC}. Currently, the start of the HL-LHC run is foreseen for mid
2026. The long shutdowns, LS2 and LS3, will be used to upgrade both the
accelerator and the detector hardware.}
    \label{fig:hl-lhc}
\end{figure}

The study of neutrinos, their mass and oscillations, can also shed light
on matter-antimatter asymmetry. The DUNE experiment will provide a huge
improvement in our ability to probe neutrino physics, detecting
neutrinos from the Long Baseline Neutrino Facility at Fermilab, as well
as linking to astro-particle physics programmes, in particular through
the potential detection of supernovas and relic neutrinos. An overview
of the experimental programme scheduled at the Fermilab facility is
given in Figure~\ref{fig:fnal-if}.

\begin{figure}
\vspace*{0.3cm}
    \centering
    \includegraphics[width=0.94\textwidth]{10yr-PLAN-9Feb-2017-V7a.pdf}
    \caption{Run schedule for the Fermilab facility until 2026.}
    \label{fig:fnal-if}
\end{figure}

In the study of the early universe immediately after the Big Bang, it is
critical to understand the phase transition between the highly
compressed quark-gluon plasma and the nuclear matter in the universe
today. The ALICE experiment at the LHC~\cite{ALICE} and
the CBM~\cite{CBM} and PANDA~\cite{PANDA} experiments at the Facility for Antiproton and Ion
Research (FAIR) are specifically designed to probe this aspect of
nuclear and particle physics. In addition ATLAS, CMS and LHCb all
contribute to the LHC heavy-ion programme.

These experimental programmes require large investments in detector
hardware, either to build new facilities and experiments (e.g., FAIR and DUNE) or to upgrade existing ones (HL-LHC, Belle II). Similarly, they
require commensurate investment in the research and development
necessary to deploy software to acquire, manage, process, and analyse
the data recorded.

For the HL-LHC, which is scheduled to begin taking data in 2026 (Figure~\ref{fig:hl-lhc}) and to run into the 2030s, some 30 times more data than
the LHC has currently produced will be collected by ATLAS and CMS. As
the total amount of LHC data already collected is close to an exabyte,
it is clear that the problems to be solved require approaches beyond
simply scaling current solutions, assuming Moore's Law and more or less
constant operational budgets. The nature of computing hardware
(processors, storage, networks) is evolving with radically new
paradigms, the quantity of data to be processed is increasing
dramatically, its complexity is increasing, and more sophisticated
analyses will be required to maximise physics yield. Developing and
deploying sustainable software for future and upgraded experiments,
given these constraints, is both a technical and a social challenge, as
detailed in this paper. An important message of this report is that a
``software upgrade'' is needed to run in parallel with the hardware
upgrades planned for the HL-LHC in order to take full advantage of these hardware upgrades and to complete the HL-LHC physics programme.

In planning for the HL-LHC in particular, it is critical that all of the
collaborating stakeholders agree on the software goals and priorities,
and that the efforts complement each other. In this spirit, the HEP
Software Foundation (HSF) began a planning exercise in late 2016 to
prepare a Community White Paper (CWP)~\cite{HSF2017} at the behest of the Worldwide LHC
Computing Grid (WLCG) project~\cite{WLCG2016}. The role of the HSF is to
facilitate coordination and common efforts in HEP software and computing
internationally and to provide a structure for the community to set
goals and priorities for future work. The objective of the CWP is to
provide a roadmap for software R\&D in preparation for the HL-LHC and
for other HEP experiments on a similar timescale, which would identify
and prioritise the software research and development investments
required:

\begin{itemize}
\item
  to achieve improvements in software efficiency, scalability and
  performance, and to make use of advances in CPU, storage and network
  technologies in order to cope with the challenges ahead;
\item
  to enable new approaches to computing and software that can radically
  extend the physics reach of the detectors;
\item
  to ensure the long-term sustainability of the software through the
  lifetime of the HL- LHC;
\item
  to ensure data and knowledge preservation beyond the lifetime of
  individual experiments;
\item
  to attract the required new expertise by offering appropriate career
  recognition to physicists specialising in software development, and by
  an effective training effort to target all contributors in the
  community.
\end{itemize}

The CWP process, organised by the HSF with the participation of the LHC
experiments and the wider HEP software and computing community, began
with a kick-off workshop at the San Diego Supercomputer Centre (SDSC),
USA, in January 2017 and concluded after a final workshop in June 2017 at
the Laboratoire d'Annecy de Physique des Particules (LAPP), France, with
a large number of intermediate topical workshops and meetings
(Appendix \ref{appendix-a---list-of-workshops}). The
entire CWP process involved an estimated 250 participants.

To reach more widely than the LHC experiments, specific contact was made
with individuals with software and computing responsibilities in the
Fermilab muon and neutrino experiments, Belle II, the Linear Collider
community, as well as various national computing organisations. The CWP
process was able to build on all the links established since the
inception of the HSF in 2014.

Working groups were established on various topics which were expected to
be important parts of the HL-LHC roadmap: \emph{Careers, Staffing and
Training; Conditions Database; Data Organisation, Management and Access;
Data Analysis and Interpretation; Data and Software Preservation;
Detector Simulation; Data-Flow Processing Frameworks; Facilities and
Distributed Computing; Machine Learning; Physics Generators; Security;
Software Development, Deployment and Validation/Verification; Software
Trigger and Event Reconstruction;} and \emph{Visualisation.} The work of
each working group is summarised in this document.

This document is the result of the CWP process. Investing in the roadmap
outlined here will be fruitful for the whole of the HEP programme and
may also benefit other projects with similar technical challenges,
particularly in astrophysics, e.g., the Square Kilometre Array (SKA)~\cite{SKA},
the Cherenkov Telescope Array (CTA)~\cite{CTA} and the Large Synoptic Survey
Telescope (LSST)~\cite{LSST}.

\hypertarget{software-and-computing-challenges}{%
\section{Software and Computing
Challenges}\label{software-and-computing-challenges}}

Run 2 for the LHC started in 2015 and delivered a proton-proton
collision energy of
\href{http://home.cern/about/engineering/restarting-lhc-why-13-tev}{13
TeV}. By the end of LHC Run 2 in 2018, it is expected that about 150
fb\textsuperscript{-1} of physics data will have been collected by both
ATLAS and CMS. Together with ALICE and LHCb, the total size of LHC data
storage pledged by sites for the year 2017 is around 1 exabyte, as shown
in Table~\ref{tab:crsg2017} from the LHC's Computing Resource Scrutiny Group
(CRSG)~\cite{Lucchesi:2284575}. The CPU allocation from the CRSG for 2017 to each
experiment is also shown.

\begin{table}
    \centering
\begin{tabular}{lp{0.15\textwidth}p{0.15\textwidth}p{0.15\textwidth}p{0.15\textwidth}}
\hline
\textbf{Experiment} & \textbf{2017 Disk Pledges (PB)} & \textbf{2017 Tape Pledges (PB)} & \textbf{Total Disk and Tape Pledges (PB)} & \textbf{2017 CPU Pledges (kHS06)}\\
\hline
ALICE & 67 & 68 & 138 & 807\\
ATLAS & 172 & 251 & 423 & 2194\\
CMS & 123 & 204 & 327 & 1729\\
LHCb & 35 & 67 & 102 & 413\\
\textbf{Total} & \textbf{400} & \textbf{591} & \textbf{990} &
\textbf{5143}\\
\hline
\end{tabular}
    \caption{Resources pledged by WLCG sites to the 4 LHC experiments for
the year 2017 as described at the September 2017 session of the
Computing Resources Scrutiny Group (CRSG).}
    \label{tab:crsg2017}
\end{table}

Using an approximate conversion from HS06~\cite{HS06} to CPU cores of 10
means that LHC computing in 2017 is supported by about 500k CPU cores.
These resources are deployed ubiquitously, from close to the experiments
themselves at CERN to a worldwide distributed computing infrastructure,
the WLCG~\cite{WLCG}. Each experiment has developed its own workflow management and
data management software to manage its share of WLCG resources.

In order to process the data, the 4 largest LHC experiments have written
more than 20 million lines of program code over the last 15 years. This
has involved contributions from thousands of physicists and many
computing professionals, encompassing a wide range of skills and
abilities. The majority of this code was written for a single
architecture (x86\_64) and with a serial processing model in mind. There
is considerable anxiety in the experiments that much of this software is
not sustainable, with the original authors no longer in the field and
much of the code itself in a poorly maintained state, ill-documented, and
lacking tests. This code, which is largely experiment-specific, manages
the entire experiment data flow, including data acquisition, high-level
triggering, calibration and alignment, simulation, reconstruction (of
both real and simulated data), visualisation, and final data analysis.

HEP experiments are typically served with a large set of integrated and
configured common software components, which have been developed either
in-house or externally. Well-known examples include ROOT~\cite{Brun1996}, which is a data analysis toolkit that
also plays a critical role in the implementation of experiments' data
storage systems, and Geant4~\cite{Agostinelli2003}, a
simulation framework through which most detector simulation is achieved.
Other packages provide tools for supporting the development process;
they include compilers and scripting languages, as well as tools for
integrating, building, testing, and generating documentation. Physics
simulation is supported by a wide range of event generators provided by
the theory community (PYTHIA~\cite{PYTHIA}, SHERPA~\cite{Gleisberg:2008ta}, ALPGEN~\cite{Mangano:565290}, MADGRAPH~\cite{MADGRAPH}, HERWIG~\cite{HERWIG},
amongst many others). There is also code developed to support the
computing infrastructure itself, such as the CVMFS distributed caching
filesystem~\cite{1742-6596-331-4-042003}, the Frontier database caching 
mechanism~\cite{Frontier}, the XRootD file access software~\cite{XRootD} and a
number of storage systems (dCache, DPM, EOS). This list of packages is % REF?
by no means exhaustive, but illustrates the range of software employed
and its critical role in almost every aspect of the programme.

Already in Run 3 LHCb will process more than 40 times the number of
collisions that it does today, and ALICE will read out Pb-Pb collisions
continuously at 50 kHz. The upgrade to the HL-LHC for Run 4 then
produces a step change for ATLAS and CMS. The beam intensity will rise
substantially, giving bunch crossings where the number of discrete
proton-proton interactions (pileup) will rise to about 200, from about 60
today. This has important consequences for the operation of the
detectors and for the performance of the reconstruction software. The
two experiments will upgrade their trigger systems to record 5-10 times
as many events as they do today. It is anticipated that HL-LHC will
deliver about 300 fb\textsuperscript{-1} of data each year.

The steep rise in resources that are then required to manage this data
can be estimated from an extrapolation of the Run 2 computing model and
is shown in Figures~\ref{fig:cms-resources} and~\ref{fig:atlas-resources}.

\begin{figure}
    \centering
    \begin{subfigure}[b]{0.47\textwidth}
        \includegraphics[width=\textwidth]{cms-cpuHLLHC.pdf}
        \caption{}
        \label{fig:cms-cpu}
    \end{subfigure}
    \quad
    \begin{subfigure}[b]{0.47\textwidth}
        \includegraphics[width=\textwidth]{cms-diskHLLHC.pdf}
        \caption{}
        \label{fig:cms-disk}
    \end{subfigure}
    \caption{CMS estimated CPU (\ref{fig:cms-cpu}) and disk space (\ref{fig:cms-disk}) 
    resources required into the HL-LHC era, using the current computing model 
    with parameters projected out for the next 12 years.}
    \label{fig:cms-resources}
\end{figure}

\begin{figure}
    \centering
    \begin{subfigure}[b]{0.9\textwidth}
    	    \centering
        \includegraphics[width=0.6\textwidth]{atlas-cpuHLLHC.pdf}
        \caption{Estimated CPU resources (in kHS06) needed for the years 2018 to 2028 for both data and simulation processing. The blue points are estimates based on the current software performance estimates and using the ATLAS computing model parameters from 2017. The solid line shows the amount of resources expected to be available if a flat funding scenario is assumed, which implies an increase of 20\% per year, based on the current technology trends.}
        \label{fig:atlas-cpu}
    \end{subfigure}
    \quad
    \begin{subfigure}[b]{0.9\textwidth}
    	    \centering
        \includegraphics[width=0.6\textwidth]{atlas-diskHLLHC.pdf}
        \caption{Estimated total disk resources (in PB) needed for the years 2018 to 2028 for both data and simulation processing. The blue points are estimates based on the current event sizes estimates and using the ATLAS computing model parameters from 2017. The solid line shows the amount of resources expected to be available if a flat funding scenario is assumed, which implies an increase of 15\% per year, based on the current technology trends.}
        \label{fig:atlas-disk}
    \end{subfigure}
    \caption{ATLAS resources required into the HL-LHC era, using the current computing model 
    and software performance.\cite{ATLAS-HL-LHC-Plots}}
    \label{fig:atlas-resources}
\end{figure}

In general, it can be said that the amount of data that experiments can
collect and process in the future will be limited by affordable software
and computing, and therefore the physics reach during HL-LHC will be
limited by how efficiently these resources can be used.

The ATLAS numbers, in Figure~\ref{fig:atlas-resources}, are particularly interesting as they
estimate the resources that will be available to the experiment if a
flat funding profile is maintained, taking into account the expected
technology improvements given current trends~\cite{Panzer2017}. As can be
seen, the shortfall between needs and bare technology gains is
considerable: a factor 4 in CPU and a factor 7 in disk in 2027.

While the density of transistors on silicon continues to increase
following Moore's Law (albeit more slowly than in the past), power
density constraints have limited the clock speed of processors for more
than a decade. This has effectively stalled any progress in the
processing capacity of a single CPU core. Instead, increases in
potential processing capacity come from increases in the core count of
CPUs and wide CPU registers. Alternative processing architectures have
become more commonplace. These range from the many-core architecture
based on standard x86\_64 cores to numerous alternatives such as GPUs.
For GPUs, the processing model is very different, allowing a much
greater fraction of the die to be dedicated to arithmetic calculations,
but at a price in programming difficulty and memory handling for the
developer that tends to be specific to each processor generation.
Further developments may even see the use of FPGAs for more
general-purpose tasks. Fully exploiting these evolutions requires a
shift in programming model to one based on \emph{concurrency}.

Even with the throttling of clock speed to limit power consumption,
power remains a major issue. Low power architectures are in huge demand.
At one level this might challenge the dominance of x86\_64 by simply
replacing it with, for example, AArch64 devices that may achieve lower power
costs for the scale of HEP computing needs than Intel has achieved with its
Xeon architecture. More extreme is an architecture that would see
specialised processing units dedicated to particular tasks, but with
possibly large parts of the device switched off most of the time,
so-called dark silicon.

Limitations in affordable storage also pose a major challenge, as does
the I/O rates of higher capacity hard disks. Network
bandwidth will probably continue to increase at the required level, but
the ability to use it efficiently will need a closer integration with
applications. This will require software developments to
support distributed computing (data and workload management, software
distribution and data access) and an increasing awareness of the
extremely hierarchical view of data, from long latency tape access and
medium-latency network access through to the CPU memory hierarchy.

Taking advantage of these new architectures and programming paradigms
will be critical for HEP to increase the ability of our code to deliver
physics results efficiently, and to meet the processing challenges of
the future. Some of this work will be focused on re-optimised
implementations of existing algorithms. This will be complicated by the
fact that much of our code is written for the much simpler model of
serial processing, and without the software engineering needed for
sustainability. Proper support for taking advantage of concurrent
programming techniques, such as vectorisation and thread-based
programming, through frameworks and libraries, will be essential, as the
majority of the code will still be written by physicists. Other
approaches should examine new algorithms and techniques, including
highly parallelised code that can run on GPUs or the use of machine
learning techniques to replace computationally expensive pieces of
simulation or pattern recognition. The ensemble of computing work that
is needed by the experiments must remain sufficiently flexible to take
advantage of different architectures that will provide computing to HEP
in the future. The use of high performance computing sites and
commercial cloud providers will very likely be a requirement for the
community and will bring particular constraints and demand flexibility.

These technical challenges are accompanied by significant human
challenges. Software is written by many people in the collaborations,
with varying levels of expertise, from a few experts with precious
skills to novice coders. This implies organising training in effective
coding techniques and providing excellent documentation, examples and
support. Although it is inevitable that some developments will remain
within the scope of a single experiment, tackling software problems
coherently as a community will be critical to achieving success in the
future. This will range from sharing knowledge of techniques and best
practice to establishing common libraries and projects that will provide
generic solutions to the community. Writing code that supports a wider
subset of the community than just a single experiment will almost
certainly be mandated upon HEP and presents a greater challenge, but the
potential benefits are huge. Attracting, and retaining, people with the
required skills who can provide leadership is another significant
challenge, since it impacts on the need to give adequate recognition to
physicists who specialise in software development. This is an important
issue that is treated in more detail later in the report.

Particle physics is no longer alone in facing these massive data
challenges. Experiments in other fields, from astronomy to genomics,
will produce huge amounts of data in the future, and will need to
overcome the same challenges that we face, i.e., massive data handling and
efficient scientific programming. Establishing links with these fields
has already started. Additionally, interest from the computing science
community in solving these data challenges exists, and mutually
beneficial relationships would be possible where there are genuine
research problems that are of academic interest to that community and
provide practical solutions to ours. The efficient processing of massive
data volumes is also a challenge faced by industry, in particular the
internet economy, which developed novel and major new technologies under
the banner of \emph{Big Data} that may be applicable to our use cases.

Establishing a programme of investment in software for the HEP
community, with a view to ensuring effective and sustainable software
for the coming decades, will be essential to allow us to reap the
physics benefits of the multi-exabyte data to come. It was in recognition of
this fact that the HSF itself was set up and already works to promote
these common projects and community developments~\cite{HSF2015}.

\hypertarget{programme-of-work}{%
\section{Programme of Work}\label{programme-of-work}}

In the following we describe the programme of work being proposed for
the range of topics covered by the CWP working groups. We summarise the
main specific challenges each topic will face, describe current
practices, and propose a number of R\&D tasks that should be undertaken
in order to meet the challenges. R\&D tasks are grouped in two different
timescales: short term (by 2020, in time for the HL-LHC Computing Technical Design Reports of
ATLAS and CMS) and longer-term actions (by 2022, to be ready for testing
or deployment during LHC Run 3).

\hypertarget{physics-generators}{%
\subsection{Physics Generators}\label{physics-generators}}

\subsubsection*{Scope and Challenges}

Monte-Carlo event generators are a vital part of modern particle
physics, providing a key component of the understanding and
interpretation of experiment data. Collider experiments have a need for
theoretical QCD predictions at very high precision. Already in LHC Run
2, experimental uncertainties for many analyses are at the same level as,
or lower than, those from theory. Many analyses have irreducible
QCD-induced backgrounds, where statistical extrapolation into the signal
region can only come from theory calculations. With future experiment
and machine upgrades, as well as reanalysis of current data, measured
uncertainties will shrink even further, and this will increase the need
to reduce the corresponding errors from theory.

Increasing accuracy will compel the use of higher-order perturbation
theory generators with challenging computational demands. Generating
Monte Carlo events using leading order (LO) generators is only a small
part of the overall computing requirements for HEP experiments.
Next-to-leading order (NLO) event generation, used more during LHC Run
2, is already using significant resources. Higher accuracy
theoretical cross sections calculated at next-to-next-to-leading (NNLO),
already important in some Run 2 analyses, are not widely used
because of computational cost. By HL-LHC the use of NNLO event
generation will be more widely required, so these obstacles to their
adoption must be overcome.
Increasing the order of the generators increases greatly the complexity
of the phase space integration required to calculate the appropriate QCD
matrix elements. The difficulty of this integration arises from the need
to have sufficient coverage in a high-dimensional space (10-15
dimensions, with numerous local maxima); the appearance of negative event
weights; and the fact that many terms in the integration cancel, so that
a very high degree of accuracy of each term is required. Memory demands
for generators have generally been low and initialisation times have
been fast, but an increase in order means that memory consumption
becomes important and initialisation times can become a significant 
fraction of the job's run time.

For HEP experiments, in many cases, meaningful predictions can only be
obtained by combining higher-order perturbative calculations with parton
showers. This procedure is also needed as high-multiplicity final states
become more interesting at higher luminosities and event rates. Matching
(N)NLO fixed-order calculations to parton shower algorithms can have a
very low efficiency, and increases further the computational load needed
to generate the necessary number of particle-level events. In addition,
many of the current models for the combination of parton-level event
generators and parton shower codes are incompatible with requirements
for concurrency on modern architectures. It is a major challenge to
ensure that this software can run efficiently on next generation
hardware and software systems.

Developments in generator software are mainly done by the HEP theory
community. Theorists typically derive career recognition and advancement from
making contributions to theory itself, rather than by making improvements
to the computational efficiency of generators \emph{per se}. So,
improving the computational efficiency of event generators, and allowing
them to run effectively on resources such as high performance computing
facilities (HPCs), will mean engaging with experts in computational
optimisation who can work with the theorists who develop generators.

The challenge in the next decade is to advance the theory and practical
implementation of event generators to support the needs of future
experiments, reaching a new level of theory precision and recognising
the demands for computation and computational efficiency that this will
bring.

\subsubsection*{Current Practice}

Extensive use of LO generators and parton shower algorithms are still
made by most HEP experiments. Each experiment has its own simulation
needs, but for the LHC experiments tens of billions of generated events
are now used each year for Monte Carlo simulations. During LHC Run 2 more and
more NLO generators were used, because of their increased theoretical
precision and stability. The raw computational complexity of NLO
amplitudes, combined with many-body phase-space evaluations and the
inefficiencies of the matching process, leads to a potentially much-increased CPU
budget for physics event simulation for ATLAS and CMS.

The use of NLO generators by the experiments today is also limited because of
the way the generators are implemented, producing significant numbers of
negative event weights. This means that the total number of events the
experiments need to generate, simulate, and reconstruct can be many 
times larger for NLO than for LO samples. At the same time, the
experiments budget only a similar number of Monte Carlo simulation events as from the real data. Having large NLO samples is thus not consistent with
existing computing budgets until a different scheme is developed that
does not depend on negative event weights or produces them only at a
significantly reduced rate.

While most event generation is run on ``standard'' grid resources,
effort is ongoing to run more demanding tasks on HPC resources, e.g.,
W-boson + 5-jet events at the Argonne Mira HPC). However, scaling for
efficient running on some of the existing HPC resources is not trivial
and requires effort.

Standard HEP libraries such as LHAPDF~\cite{LHAPDF}, HepMC\cite{HepMC}, 
and Rivet~\cite{Rivet} are used by the generators for integration into
the experiments' event generation workflows. These require extensions
and sustained maintenance that should be considered a shared
responsibility of the theoretical and experimental communities in the
context of large-scale experiments. In practice, however, it has been
difficult to achieve the level of support that is really needed as there
has been a lack of recognition for this work. To help improve the
capabilities and performance of generators as used by the experimental
HEP programme, and to foster interaction between the communities, the 
MCnet~\cite{MCnet} short-term studentship programme has been very useful.
Interested experimental PhD students can join a generator group for
several months to work on improving a physics aspect of the simulation
that is relevant to their work, or to improve the integration of the
generator into an experimental framework.

\subsubsection*{Research and Development Programme}

As the Monte Carlo projects are funded mainly to develop theoretical
improvements, and not mainly as ``suppliers'' to the experimental HEP
programme, any strong requests towards efficiency improvements from the
experimental community would need to be backed up by plausible avenues
of support that can fund contributions from software engineers with the
correct technical skills in software optimisation to work within the
generator author teams.

In a similar way to the MCnet studentships, a matchmaking scheme could
focus on the software engineering side, and transfer some of the
expertise available in the experiments and facilities teams to the
generator projects. Sustainable improvements are unlikely to be
delivered by graduate students ``learning on the job'' and then leaving
after a few months, so meeting the requirement of transferring technical
expertise and effort will likely require placements for experienced
optimisation specialists and a medium- to long-term connection to the
generator project.

HEP experiments, which are now managed by very large collaborations including many
technical experts, can also play a key role in sustaining a healthy
relationship between theory and experiment software. Effort to work on
common tools that benefit both the experiment itself and the wider
community would provide shared value that justifies direct investment
from the stakeholders. This model would also be
beneficial for core HEP tools like LHAPDF, HepMC and Rivet, where future
improvements have no theoretical physics interest anymore, putting them
in a similar situation to generator performance improvements. One
structural issue blocking such a mode of operation is that some
experiments do not currently recognise contributions to external
projects as experiment service work --- a situation deserving of review
in areas where external software tools are critical to experiment
success.

In the following we describe specific areas of R\&D for event generation 
up to 2022 and beyond.

\begin{itemize}
\item
  The development of new and improved theoretical algorithms provides
  the largest potential for improving event generators. While it is not
  guaranteed that simply increasing the effort dedicated to this task
  will bring about the desired result, the long-term support of event
  generator development, and the creation of career opportunities in
  this research area, are critical given the commitment to experiments
  on multi-decade scales.
\item
  Expand development in reweighting event samples, where new physics
  signatures can be explored by updating the partonic weights according
  to new matrix elements. It is necessary that the phase space for the
  updated model be a subset of the original one, which is an important
  limitation. The procedure is more complex at NLO and can require
  additional information to be stored in the event files to properly
  reweight in different cases. Overcoming the technical issues from
  utilising negative event weights is crucial. Nevertheless, the method
  can be powerful in many cases, and would hugely reduce the time needed
  for the generation of BSM samples.
\item
  At a more technical level, concurrency is an avenue that has yet to be
  explored in depth for event generation. As the calculation of matrix
  elements requires VEGAS-style integration, this work would be helped
  by the development of a new Monte-Carlo integrator. For multi-particle
  interactions, factorising the full phase space integration into lower
  dimensional integrals would be a powerful method of parallelising,
  while the interference between different Feynman graphs can be handled
  with known techniques.
\item
  For many widely used generators, basic problems of concurrency and
  thread hostility need to be tackled, to make these packages suitable
  for efficient large scale use on modern processors and within modern
  HEP software frameworks. Providing appropriate common tools for
  interfacing, benchmarking and optimising multithreaded code would
  allow expertise to be shared effectively~\cite{HSF-CWP-2017-13}.
\item
  In most generators, parallelism was added post-facto, which leads to
  scaling problems when the level of parallelism becomes very large,
  e.g., on HPC machines. These HPC machines will be part of the
  computing resource pool used by HEP, so solving scaling issues on
  these resources for event generation is important, particularly as the
  smaller generator code bases can make porting to non-x86\_64
  architectures more tractable. The problem of long and inefficient
  initialisation when a job utilises hundreds or thousands of cores on
  an HPC needs to be tackled. While the memory consumption of event
  generators is generally modest, the generation of tree-level
  contributions to high multiplicity final states can use significant
  memory, and gains would be expected from optimising here.
\item
  An underexplored avenue is the efficiency of event generation as
  used by the experiments. An increasingly common usage is to generate
  very large inclusive event samples, which are filtered on event
  final-state criteria to decide which events are to be retained and
  passed onto detector simulation and reconstruction. This naturally
  introduces a large waste of very CPU-expensive event generation, which
  could be reduced by developing filtering tools within the generators
  themselves, designed for compatibility with the experiments'
  requirements. A particularly wasteful example is where events are
  separated into orthogonal subsamples by filtering, in which case the
  same large inclusive sample is generated many times, with each stream
  filtering the events into a different group: allowing a single
  inclusive event generation to be filtered into several orthogonal
  output streams would improve efficiency.
\end{itemize}

\hypertarget{detector-simulation}{%
\subsection{Detector Simulation}\label{detector-simulation}}

\subsubsection*{Scope and Challenges}

For all its success so far, the challenges faced by the HEP field in the
simulation domain are daunting. During the first two runs, the LHC
experiments produced, reconstructed, stored, transferred, and analysed
tens of billions of simulated events. This effort required more than
half of the total computing resources allocated to the experiments. As
part of the HL-LHC physics programme, the upgraded
experiments expect to collect 150 times more data than in Run 1; demand
for larger simulation samples to satisfy analysis needs will grow
accordingly. In addition, simulation tools have to serve diverse
communities, including accelerator-based particle physics research
utilising proton-proton colliders, neutrino, dark matter, and muon
experiments, as well as the cosmic frontier. The complex detectors of
the future, with different module- or cell-level shapes, finer
segmentation, and novel materials and detection techniques, require
additional features in geometry tools and bring new demands on physics
coverage and accuracy within the constraints of the available computing
budget. The diversification of the physics programmes also requires new
and improved physics models. More extensive use of Fast Simulation is a
potential solution, under the assumption that it is possible to improve
time performance without an unacceptable loss of physics accuracy.

The gains that can be made by speeding up critical elements of the
Geant4 simulation toolkit can be leveraged for all applications that use
it, and it is therefore well worth the investment in effort needed to
achieve it. The main challenges to be addressed if the required physics
and software performance goals are to be achieved are:

\begin{itemize}
\item
  Reviewing the physics models' assumptions, approximations, and
  limitations in order to achieve higher precision, and to extend the
  validity of models up to energies of the order of 100 TeV foreseen
  with the Future Circular Collider (FCC) project~\cite{FCC}.
\item
  Redesigning, developing, and commissioning detector simulation
  toolkits to be more efficient when executed on current vector CPUs and
  emerging new architectures, including GPUs, where use of SIMD
  vectorisation is vital; this includes porting and optimising the
  experiments' simulation applications to allow exploitation of large
  HPC facilities.
\item
  Exploring different Fast Simulation options, where the full detector
  simulation is replaced, in whole or in part, by computationally
  efficient techniques. An area of investigation is common frameworks
  for fast tuning and validation.
\item
  Developing, improving and optimising geometry tools that can be shared
  am\-ong experiments to make the modeling of complex detectors
  computationally more efficient, modular, and transparent.
\item
  Developing techniques for background modeling, including contributions
  of multiple hard interactions overlapping the event of interest in
  collider experiments (pileup).
\item
  Revisiting digitisation algorithms to improve performance and
  exploring opportunities for code sharing among experiments.
\item
  Recruiting, training, retaining human resources in all areas of
  expertise pertaining to the simulation domain, including software and
  physics.
\end{itemize}

It is obviously of critical importance that the whole community of
scientists working in the simulation domain continue to work together in
as efficient a way as possible in order to deliver the required
improvements. Very specific expertise is required across all simulation
domains, such as physics modeling, tracking through complex geometries
and magnetic fields, and building realistic applications that accurately
simulate highly complex detectors. Continuous support is needed to
recruit, train, and retain people with a unique set of skills needed to
guarantee the development, maintenance, and support of simulation codes
over the long timeframes foreseen in the HEP experimental programme.

\subsubsection*{Current Practices}

The Geant4 detector simulation toolkit is at the core of simulation in
almost every HEP experiment. Its continuous development, maintenance,
and support for the experiments is of vital importance. New or refined
functionality in physics coverage and accuracy continues to be delivered
in the ongoing development programme and software
performance improvements are introduced whenever possible.

>>>>

\begin{appendices}
